A comparison of design-based and model-based approaches for spatial data.

### 2021-06-09 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
adjusted grts section and added simulation one table

@@

<!-- \citep{Cressie2015statistics, Schabenberger2017statistical} -->

@@

Assuming the data is a realization of a specific data-generating process yields predictors that are linked to distributional assumptions. These distributional assumptions are used to derive prediction intervals. The distributional assumptions allow the prediction intervals can be more precise. @Cressie2015statistics and @Schabenberger2017statistical provide reviews of model-based approaches for spatial data.

@@

```{r, echo = FALSE, message = FALSE, warning = FALSE, out.width = "100%"}
library(tidyverse)
library(here)

set.seed(06092021)

@@

Spatially balanced sampling algorithms use spatial information to obtain samples spread out in space. Spatially balanced samples are useful because they tend to yield estimators that are more precise than estimators constructed from a sampling algorithm that is not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. To quantify spatial balance, @stevens2004spatially proposed statistics based on Voroni polygons. Many spatially balanced sampling algorithms exist, including the Generalized Random Tessellation Stratified [@stevens2004spatially], the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. Here we focus on the Generalized Random Tessellation Stratified (GRTS) algorithm, which has several attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample frames. A finite sample frame contains a finite number of sampling units and is related to a point geometry. An infinite sample frame contains an infinite number of sampling units and is related to linear and polygon geometries. Examples of point, linear, and polygon resources include lake centroids, stream networks, and wetland areas, respectively. In addition to its applicability for finite and infinite sample frames, the GRTS algorithm naturally accommodates stratified designs and designs with unequal selection probabilities. The algorithm has also been used to select replacement sites using reverse hierarchical ordering [@stevens2004spatially]. Replacement sites are used to replace sites in the original sample that cannot be sampled, often as a result of physical difficulty in reaching the site or landowner denial of access to the sites. More recently, the GRTS algorithm also accommodates legacy (historical) sites, minimum distance between sites, and nearest neighbor replacement sites. The GRTS algorithm is implemented in the \textbf{\textsf{R}} package \texttt{spsurvey} [@dumelle2021spsurvey].

The GRTS algorithm works by mapping two-dimensional space into one-dimensional space while incorporating some elements of randomness. First a square bounding box is placed over the sample frame and divided into four equally sized squares randomly given an level-one address of 0, 1, 2, or 3. Then the inclusion probabilities of the sampling units in each square are summed. If this sum is at least one in any of the cells, a set of four sub-squares are placed within each square. Each set of four sub-squares are randomly given a second address (a level-two address) of 0, 1, 2, 3. Then the inclusion probabilities in each sub-square are summed, and if any of these sums are at least one, a third level of squares are added in the same manner. This hierarchical addressing continues until the inclusion probabilities in the smallest squares are all less than 1. The squares are then mapped to the one-dimensional line in order of their addresses. The length of this line equals $n$, the desired sample size (the sum of the inclusion probabilities). A uniform random variable, denoted by $u_1$ is simulated in $[0, 1]$ and placed on this line. The location of $u_1$ on the line can be linked to a sampling unit, denoted by $s_1$. Then $u_2 \equiv u_1 + 1$, which can be linked to another sampling unit, denoted by $s_2$. Because the GRTS algorithm requires the inclusion probabilities of the smallest sub-cells to be less than one, $s_1$ and $s_2$ must be distinct. This process continues until the $n$ sampling units have been selected as part of the sample.  Further details are provided by [@stevens2004spatially] and [@dumelle2021spsurvey].

The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population totals (and means). If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$, denoted by $\hat{\tau}_{HT}$, is given by
\begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the observed value of the $i$th sampling unit, and $\pi_i$ is the inclusion probability of the $i$th sampling unit. Estimating the variance of $\hat{\tau}_{HT}$ directly relies on knowing the probability that $s_i$ and $s_j$ are both included in the sample (second-order inclusion probability), which can be unfeasible to compute. Furthermore, this variance estimator does not incorporate spatial locations. To address these challenges, [@stevens2003variance] proposed an alternative variance estimator: the local neighborhood variance estimator. The local neighborhood variance estimator does not require second-order inclusion probabilities. It also incorporates spatial locations, which tends to reduce the estimated variance of compared $\hat{\tau}_{HT}$ to a variance estimator ignoring spatial location [@stevens2003variance].

@@

cases

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE,
  include = TRUE, 
  echo = FALSE
)
library(tidyverse)
library(here)
library(xtable) # latex tables
```

@@

@@

Assuming the data is a realization of a specific data-generating process yields predictors that are linked to distributional assumptions. These distributional assumptions are used to derive prediction intervals. The distributional assumptions allow the prediction intervals can be more precise. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data.

@@

```{r, out.width = "85%"}
set.seed(06092021)

@@

Sampling designs can incorporate spatial locations to obtain samples that are spatially balanced with respect to the population [@stevens2004spatially]. A sample is spatially with respect to the population if the sampled population units are a miniature of the population units. A sample is a miniature of the population if the distribution of the sampled population units mirrors the density of all population units. Spatial balance with respect to the population is is different than spatial balance with respect to geography. A sample that is spatially balanced with respect to geography is spread out in some type of equidistant manner over geographical space and are not meant to be miniatures of the population. When we refer to spatial balance henceforth, we mean spatial balance with respect to the population.

Spatially balanced samples are useful because they tend to yield estimates that have lower variance than estimates constructed from sampling designs lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. To quantify spatial balance, @stevens2004spatially proposed loss functions based on Voroni polygons. The first spatiallly balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified [@stevens2004spatially]. Since GRTS was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. We focus on the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced sampling because the algorithm has several attractive properties detailed by @stevens2004spatially and @dumelle2021spsurvey.

The GRTS algorithm is used to sample from finite and infinite populations and works by utilizing a mapping between two-dimensional and one-dimensional space. The population units in two-dimensional space are divided into cells using a hierarchical index. Population units are then mapped to a one-dimensional line via the hierarchical indexing. The line length of each population unit equals its inclusion probability. A systematic sample is conducted on the line and these samples are linked to a population unit in two-dimensional space, which results in the desired sample size. @stevens2004spatially provide and @dumelle2021spsurvey provide further details. The GRTS algorithm is available in the \textbf{\textsf{R}} package \texttt{spsurvey} [@dumelle2021spsurvey].

After collecting a sample, the data are used to estimate population parameters. The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population means and totals. If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$ (denoted by $\hat{\tau}_{ht}$), is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ and $pi_i$ are the observed value and inclusion probability of the $i$th population unit selected in the sample. @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but they have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and population unit $j$ are included in the sample, which can be very difficult to calculate. Second, they ignore the spatial locations of the population units. To address these drawbacks, @stevens2003variance proposed a local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$, and it incorporates spatial locations by assigning higher weights to nearby observations. @stevens2003variance show this variance estimators tends to reduce the variability associated with estimating $\tau$. This yields confidence intervals for $\tau$ that are narrower than confidence intervals constructed from variance estimators ignoring spatial locations.

@@

```{r, results = "asis"}
sim_one <- read_csv(here("inst", "output", "sim_one", "summ_output.csv"))
colnames(sim_one) <- c("Approach", "Bias", "RMSE", "MedAE", "Coverage", "PClose", "MedIL")
sim_one_table <- xtable(sim_one, digits = 4, caption = "Approach, mean bias (Bias), root-mean-squared error (RMSE), median absolute error (MedAE), 95 percent interval coverage (Coverage), proportion of times the approach was closer to the true value (PClose), and median interval length (MedIL)", type = "latex", latex.environments = "center", label = "tab:sim_one")
print(sim_one_table, include.rownames = FALSE, comment = FALSE)
```

### 2021-06-09 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
adjusted grts section and added simulation one table

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as variables measured at specific
geographic locations. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered, but
they tend to compare design-based approaches that ignore spatial
location in sampling to model-based approaches (Brus and De Gruijter,
1997; Ver Hoef, 2002; Ver Hoef, 2008). More recent overviews include
Brus (2020) and Wang et al. (2012), but no numerical comparison has been
made between design-based approaches that incorporate spatial location
in sampling and model-based approaches.

@@

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals can be more
precise. (\textbf{Cressie2015statistics?}) and
(\textbf{Schabenberger2017statistical?}) provide reviews of model-based
approaches for spatial data.

Several comparisons between the model-based paradigm and the
design-based paradigm have been made in many different contexts. Sterba
(2009) give some history of the two paradigms as well as some
applications in psychological contexts. Cooper (2006) review the two
approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g. Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

\includegraphics[width=1\linewidth]{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-1-1}

@@

Spatially balanced sampling algorithms use spatial information to obtain
samples spread out in space. Spatially balanced samples are useful
because they tend to yield estimators that are more precise than
estimators constructed from a sampling algorithm that is not spatially
balanced (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens Jr and
Olsen, 2004; Wang et al., 2013). To quantify spatial balance, Stevens Jr
and Olsen (2004) proposed statistics based on Voroni polygons. Many
spatially balanced sampling algorithms exist, including the Generalized
Random Tessellation Stratified (Stevens Jr and Olsen, 2004), the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
(Benedetti and Piersimoni, 2017), and Halton Iterative Partitioning
(Robertson et al., 2018) algorithms. Here we focus on the Generalized
Random Tessellation Stratified (GRTS) algorithm, which has several
attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample
frames. A finite sample frame contains a finite number of sampling units
and is related to a point geometry. An infinite sample frame contains an
infinite number of sampling units and is related to linear and polygon
geometries. Examples of point, linear, and polygon resources include
lake centroids, stream networks, and wetland areas, respectively. In
addition to its applicability for finite and infinite sample frames, the
GRTS algorithm naturally accommodates stratified designs and designs
with unequal selection probabilities. The algorithm has also been used
to select replacement sites using reverse hierarchical ordering (Stevens
Jr and Olsen, 2004). Replacement sites are used to replace sites in the
original sample that cannot be sampled, often as a result of physical
difficulty in reaching the site or landowner denial of access to the
sites. More recently, the GRTS algorithm also accommodates legacy
(historical) sites, minimum distance between sites, and nearest neighbor
replacement sites. The GRTS algorithm is implemented in the
\textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle et al., 2021).

The GRTS algorithm works by mapping two-dimensional space into
one-dimensional space while incorporating some elements of randomness.
First a square bounding box is placed over the sample frame and divided
into four equally sized squares randomly given an level-one address of
0, 1, 2, or 3. Then the inclusion probabilities of the sampling units in
each square are summed. If this sum is at least one in any of the cells,
a set of four sub-squares are placed within each square. Each set of
four sub-squares are randomly given a second address (a level-two
address) of 0, 1, 2, 3. Then the inclusion probabilities in each
sub-square are summed, and if any of these sums are at least one, a
third level of squares are added in the same manner. This hierarchical
addressing continues until the inclusion probabilities in the smallest
squares are all less than 1. The squares are then mapped to the
one-dimensional line in order of their addresses. The length of this
line equals \(n\), the desired sample size (the sum of the inclusion
probabilities). A uniform random variable, denoted by \(u_1\) is
simulated in \([0, 1]\) and placed on this line. The location of \(u_1\)
on the line can be linked to a sampling unit, denoted by \(s_1\). Then
\(u_2 \equiv u_1 + 1\), which can be linked to another sampling unit,
denoted by \(s_2\). Because the GRTS algorithm requires the inclusion
probabilities of the smallest sub-cells to be less than one, \(s_1\) and
\(s_2\) must be distinct. This process continues until the \(n\)
sampling units have been selected as part of the sample. Further details
are provided by (Stevens Jr and Olsen, 2004) and (Dumelle et al., 2021).

The Horvitz-Thompson estimator (Horvitz and Thompson, 1952) and its
continuous analog (Cordy, 1993) yield unbiased estimates of population
totals (and means). If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\), denoted by \(\hat{\tau}_{HT}\),
is given by \begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the observed value of the \(i\)th sampling
unit, and \(\pi_i\) is the inclusion probability of the \(i\)th sampling
unit. Estimating the variance of \(\hat{\tau}_{HT}\) directly relies on
knowing the probability that \(s_i\) and \(s_j\) are both included in
the sample (second-order inclusion probability), which can be unfeasible
to compute. Furthermore, this variance estimator does not incorporate
spatial locations. To address these challenges, (Stevens Jr and Olsen,
2003) proposed an alternative variance estimator: the local neighborhood
variance estimator. The local neighborhood variance estimator does not
require second-order inclusion probabilities. It also incorporates
spatial locations, which tends to reduce the estimated variance of
compared \(\hat{\tau}_{HT}\) to a variance estimator ignoring spatial
location (Stevens Jr and Olsen, 2003).

@@

cases

@@

@@

>>>>

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as variables measured at specific
geographic locations. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered, but
they tend to compare design-based approaches that ignore spatial
location in sampling to model-based approaches (Brus and De Gruijter,
1997; Ver Hoef, 2002, 2008). More recent overviews include Brus (2020)
and Wang et al. (2012), but no numerical comparison has been made
between design-based approaches that incorporate spatial location in
sampling and model-based approaches.

@@

Assuming the data is a realization of a specific data-generating process
yields predictors that are linked to distributional assumptions. These
distributional assumptions are used to derive prediction intervals. The
distributional assumptions allow the prediction intervals can be more
precise. Cressie (1993) and Schabenberger and Gotway (2017) provide
reviews of model-based approaches for spatial data.

Several comparisons between the model-based paradigm and the
design-based paradigm have been made in many different contexts. Sterba
(2009) give some history of the two paradigms as well as some
applications in psychological contexts. Cooper (2006) review the two
approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g.~Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

@@

\includegraphics[width=0.85\linewidth]{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-2-1}

@@

Sampling designs can incorporate spatial locations to obtain samples
that are spatially balanced with respect to the population (Stevens Jr
and Olsen, 2004). A sample is spatially with respect to the population
if the sampled population units are a miniature of the population units.
A sample is a miniature of the population if the distribution of the
sampled population units mirrors the density of all population units.
Spatial balance with respect to the population is is different than
spatial balance with respect to geography. A sample that is spatially
balanced with respect to geography is spread out in some type of
equidistant manner over geographical space and are not meant to be
miniatures of the population. When we refer to spatial balance
henceforth, we mean spatial balance with respect to the population.

Spatially balanced samples are useful because they tend to yield
estimates that have lower variance than estimates constructed from
sampling designs lacking spatial balance (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens Jr and Olsen, 2004; Wang et al., 2013). To quantify
spatial balance, Stevens Jr and Olsen (2004) proposed loss functions
based on Voroni polygons. The first spatiallly balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (Stevens Jr and Olsen, 2004). Since GRTS was
developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning (Robertson et al., 2018) algorithms. We
focus on the Generalized Random Tessellation Stratified (GRTS) algorithm
to select spatially balanced sampling because the algorithm has several
attractive properties detailed by Stevens Jr and Olsen (2004) and
Dumelle et al. (2021).

The GRTS algorithm is used to sample from finite and infinite
populations and works by utilizing a mapping between two-dimensional and
one-dimensional space. The population units in two-dimensional space are
divided into cells using a hierarchical index. Population units are then
mapped to a one-dimensional line via the hierarchical indexing. The line
length of each population unit equals its inclusion probability. A
systematic sample is conducted on the line and these samples are linked
to a population unit in two-dimensional space, which results in the
desired sample size. Stevens Jr and Olsen (2004) provide and Dumelle et
al. (2021) provide further details. The GRTS algorithm is available in
the \textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle et al.,
2021).

After collecting a sample, the data are used to estimate population
parameters. The Horvitz-Thompson estimator (Horvitz and Thompson, 1952)
and its continuous analog (Cordy, 1993) yield unbiased estimates of
population means and totals. If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\) (denoted by \(\hat{\tau}_{ht}\)),
is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) and \(pi_i\) are the observed value and
inclusion probability of the \(i\)th population unit selected in the
sample. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but they have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are included in the sample, which
can be very difficult to calculate. Second, they ignore the spatial
locations of the population units. To address these drawbacks, Stevens
Jr and Olsen (2003) proposed a local neighborhood variance estimator.
The local neighborhood variance estimator does not rely on \(\pi_{ij}\),
and it incorporates spatial locations by assigning higher weights to
nearby observations. Stevens Jr and Olsen (2003) show this variance
estimators tends to reduce the variability associated with estimating
\(\tau\). This yields confidence intervals for \(\tau\) that are
narrower than confidence intervals constructed from variance estimators
ignoring spatial locations.

@@

\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  \hline
Approach & Bias & RMSE & MedAE & Coverage & PClose & MedIL \\ 
  \hline
Design & 0.0003 & 0.0353 & 0.0251 & 0.9461 & 0.4889 & 0.1362 \\ 
  Model & -0.0001 & 0.0362 & 0.0253 & 0.9480 & 0.5111 & 0.1430 \\ 
   \hline
\end{tabular}
\caption{Approach, mean bias (Bias), root-mean-squared error (RMSE), median absolute error (MedAE), 95 percent interval coverage (Coverage), proportion of times the approach was closer to the true value (PClose), and median interval length (MedIL)} 
\label{tab:sim_one}
\end{table}

@@

\leavevmode\hypertarget{ref-cressie1993statistics}{}%
Cressie, N., 1993. Statistics for spatial data. John Wiley \& Sons.

@@

\leavevmode\hypertarget{ref-schabenberger2017statistical}{}%
Schabenberger, O., Gotway, C.A., 2017. Statistical methods for spatial
data analysis. CRC press.

\leavevmode\hypertarget{ref-sen1953estimate}{}%
Sen, A.R., 1953. On the estimate of the variance in sampling with
varying probabilities. Journal of the Indian Society of Agricultural
Statistics 5, 127.
